---
title: "BIG5 Selector: Data Audit and Validation EDA"
output:
  html_document:
    toc: true
params:
  project_root: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
root <- normalizePath(params$project_root, winslash = "/", mustWork = TRUE)
source(file.path(root, "R", "common.R"))
pmr_use_library(root); pmr_require(c("jsonlite", "knitr", "digest"))
rows_path <- file.path(root, "data", "processed", "selector", "rows.csv")
if (!file.exists(rows_path)) stop("Prepared selector rows.csv is required.")
rows <- pmr_read_csv(rows_path)
source(file.path(root, "R", "selector", "core.R")); selector_validate_rows(rows)
manifest_path <- file.path(root, "reports", "selector", "source_manifest.json")
cleaning_path <- file.path(root, "reports", "selector", "cleaning.json")
raw_path <- file.path(root, "data", "raw", "selector", "big5_chat_dataset.csv")
if (!all(file.exists(c(raw_path, manifest_path, cleaning_path))))
  stop("The official BIG5-CHAT CSV, source manifest, and cleaning report are required.")
manifest <- jsonlite::fromJSON(manifest_path)
cleaning <- jsonlite::fromJSON(cleaning_path)
raw_rows <- nrow(pmr_read_csv(raw_path))
raw_bytes <- file.info(raw_path)$size
raw_sha256 <- digest::digest(raw_path, algo = "sha256", file = TRUE, serialize = FALSE)
if (!identical(raw_sha256, manifest$sha256)) stop("BIG5-CHAT SHA-256 disagrees with source_manifest.json.")
if (!identical(as.numeric(raw_bytes), as.numeric(manifest$bytes))) stop("BIG5-CHAT byte size disagrees with source_manifest.json.")
if (raw_rows != cleaning$raw_rows || nrow(rows) != cleaning$kept_rows ||
    cleaning$raw_rows != cleaning$kept_rows + cleaning$excluded_rows)
  stop("Raw, kept, and excluded row counts disagree with cleaning.json.")
```

This report reads the official BIG5-CHAT source CSV itself, verifies its bytes and SHA-256 against the pinned source manifest, and then audits the Python-prepared classifier table. It does not claim that R produced the cleaned table.

## Official source audit

```{r raw-counts}
raw_table <- data.frame(
  source = "wenkai-li/big5_chat",
  revision = manifest$revision,
  observed_rows = format(raw_rows, scientific = FALSE, trim = TRUE),
  kept_rows = format(nrow(rows), scientific = FALSE, trim = TRUE),
  excluded_rows = format(cleaning$excluded_rows, scientific = FALSE, trim = TRUE),
  observed_bytes = format(raw_bytes, scientific = FALSE, trim = TRUE),
  manifest_bytes = format(manifest$bytes, scientific = FALSE, trim = TRUE),
  sha256_match = raw_sha256 == manifest$sha256
)
knitr::kable(raw_table, digits = 0,
             caption = "Official BIG5-CHAT source verified against the pinned manifest and cleaning audit")
```

## Prepared classifier table

```{r prepared-summary}
prepared <- as.data.frame(xtabs(~ split + role, rows))
knitr::kable(prepared, caption = "Prepared selector rows by split and role")
```

```{r label-balance}
labels <- as.data.frame(xtabs(~ split + label, rows))
knitr::kable(labels, caption = "BIG5 class balance by prepared split")
```

```{r group-audit}
groups <- aggregate(row_id ~ split + group_id, rows, length)
group_summary <- do.call(rbind, lapply(split(groups$row_id, groups$split), function(x)
  data.frame(groups = length(x), min_rows = min(x), median_rows = median(x), max_rows = max(x))))
group_summary$split <- rownames(group_summary); rownames(group_summary) <- NULL
knitr::kable(group_summary[c("split", "groups", "min_rows", "median_rows", "max_rows")],
             caption = "Group counts and rows per group after preparation")
```

The classifier contract requires each `group_id` to occur in exactly one split. The validation above stops rendering if this invariant, the fixed ten-class vocabulary, or unique `row_id` values fail.

## Training-text length and group size

These plots use training rows. Length means Unicode characters, not language-model tokens. The logarithmic length axis preserves the long tail without dropping records.

```{r length-audit, fig.width=9, fig.height=4}
training <- rows[rows$split == "train", ]
response_chars <- nchar(training$model_response, type = "chars")
knitr::kable(data.frame(percentile = c("minimum", "median", "90%", "99%", "maximum"),
                       characters = as.numeric(quantile(response_chars, c(0, .5, .9, .99, 1)))),
             caption = "Target response length in training data")
hist(log10(pmax(response_chars, 1)), breaks = 35, col = "#287F85", border = "white",
     main = "Training responses vary in length", xlab = "log10(response characters)", ylab = "Records")
```

```{r group-size-plot, fig.width=9, fig.height=4}
group_size <- groups$row_id[groups$split == "train"]
hist(group_size, breaks = seq(0, max(group_size) + 10, by = 10), col = "#526F9B", border = "white",
     main = "Connected groups stay intact", xlab = "Records per training group", ylab = "Groups")
```

## Validation-only model comparison

```{r validation-results}
metrics_path <- file.path(root, "artifacts", "selector", "validation_metrics.csv")
if (file.exists(metrics_path)) {
  metrics <- pmr_read_csv(metrics_path)
  metrics$validation_macro_f1 <- as.numeric(metrics$validation_macro_f1)
  knitr::kable(metrics[order(-metrics$validation_macro_f1), ], digits = 4,
               caption = "Candidate selection using validation Macro-F1 only")
} else {
  cat("Training outputs are not present yet. This section intentionally reports no test result.")
}
```

Test labels are not read or summarized as model results here. The separate evaluation entry point remains locked until a hash-pinned frozen marker exists.
